---
output: github_document
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  fig.height = 4,
  fig.width = 8,
  out.width = "100%",
  dev = 'svg'
)

set.seed(1)
```

# Process Behaviour Charts

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

An R package for run charts and individuals control charts for statistical quality control and improvement. 

pbcharts implements the I prime (I' or normalised I) control chart suggested by Taylor <https://variation.com/normalized-individuals-control-chart/>.

The I' chart adjusts control limits to varying subgroup sizes making pbcharts useful for a wide range of measurement and count data and a convenient replacement for the classic Shewhart control charts.

pbcharts uses only functions from base R making in fast and robust.

pbcharts is currently able to:

* test for special cause variation using runs analysis and control limits;
* signal special causes using clear visual clues;
* freeze calculations of centre line and control limits to a baseline period;
* split charts into separate periods;
* exclude individual data points from calculations;
* facet plots (small multiples) on one categorical variable.

pbcharts is in early develpment. Please report any issues at https://github.com/anhoej/pbcharts/issues

## Installation

You can install the development version of pbcharts from [GitHub](https://github.com/anhoej/pbcharts) with:

```
devtools::install_github("anhoej/pbcharts")
```

## Examples

Draw a run chart of 24 random normal values:

```{r}
library(pbcharts)
pbc(rnorm(24))
```

Draw an individuals (I) control chart:

```{r}
pbc(rnorm(24), chart = 'i')
```

Signal special causes from data points outside control limits (red points) and(or) unusually long or few runs (red and dashed centre line):

```{r}
pbc(1:11, chart = 'i')
```

Standard I chart (subgroup size = 1) of average decision to delivery times for grade 2 caesarian sections (C-section):

```{r}
pbc(month, avg_delay, 
    data  = csection, 
    chart = 'i')
```

I' chart of C-section data taking varying subgroup sizes into account:

```{r}
pbc(month, avg_delay * n, n,  # multiply numerator and denominator to keep scale
    data = csection,
    chart = 'i')
```

Standard I chart of average HbA1c in children with diabetes:

```{r}
pbc(month, avg_hba1c,
    data  = hba1c,
    chart = 'i',
    title = 'I chart of average HbA1c in children with diabetes',
    ylab  = 'mmol/mol',
    xlab  = 'Month')
```

I' chart of average HbA1c in children with diabetes with denominator (number of children):

```{r}
pbc(month, avg_hba1c * n, n,
    data  = hba1c,
    chart = 'i',
    title = 'I\' chart of average HbA1c in children with diabetes',
    ylab  = 'mmol/mol',
    xlab  = 'Month')
```

Structure and summary of a pbc object:

```{r}
# save pbc object while suppressing plotting
p <- pbc(month, avg_hba1c * n, n,
         data  = hba1c,
         chart = 'i',
         title = 'I\' chart of average HbA1c in children with diabetes',
         ylab  = 'mmol/mol',
         xlab  = 'Month',
         plot = FALSE)

# print structure
str(p)

# print summary
summary(p)

# plot chart
plot(p)
```


Faceted I' chart of bacteremia mortality in six hospitals:

```{r, fig.height=7}
pbc(month, deaths, cases,
    facet    = hospital,                # facet plot by hospital
    data     = bacteremia_mortality,
    chart    = 'i',
    multiply = 100,                     # show percent rather than proportions
    ylim     = c(0, NA),                # suppress negative control limits
    title    = 'Bacteremia mortality',
    ylab     = '%',
    xlab     = 'Month')
```

Print a summary:

```{r}
p <- pbc(month, deaths, cases, hospital,
         data  = bacteremia_mortality,
         chart = 'i',
         plot  = FALSE)

summary(p)
```

Plot a pbc object:

```{r, fig.height=7}
plot(p)
```

## Procedure for calculating centre line and conrol limits

We use the following symbols:

* n = numerators
* d = denominators
* o = number of data values
* i = i^th^ data value

Values to plot:

$$
y = \frac{n}{d}
$$

Centre line:

$$
CL = \frac{\sum{n}}{\sum{d}}
$$

Standard deviation of i^th^ data point:

$$
s_i = \sqrt{\frac{\pi}{2}}\frac{\vert{}y_i-y_{i-1}\vert{}}{\sqrt{\frac{1}{d_i}+\frac{1}{d_{i-1}}}}
$$

Average standard deviation:

$$
\bar{s} = \frac{\sum{s}}{o}
$$

Control limits:

$$
\text{control limits} = CL \pm 3 \frac{\bar{s}}{\sqrt{d_i}}
$$

## Tests for special cause variation

pbc employs three tests for special cause variation:

* **Data points outside the control limits**.

* **Unusually long runs**: A run is one or more consecutive data points on the same side of the centre line. Data points that fall directly on the centre line neither break nor contribute to the run. The upper 95% prediction limit for longest run in a random process is approximately $log_2(n)+3$ (rounded to the nearest integer), where $n$ is the number of useful data points (data points not on the centre line).

* **Unusually few crossings**: A crossing is when two consecutive data points are on opposite sides of the centre line. In a random process, the number of crossings follows a binomial distribution. The lower 5% prediction limit for number of crossings is found using the cumulative probability distribution, `qbinom(p = 0.05, size = n - 1, prob = 0.5)`.

Critical values for runs and crossing:

```{r, results='asis', echo=FALSE}
n <- 10:100
x <- data.frame(
  `Number of useful observations`       = n,
  `Upper limit for longest run`         = round(log2(n) + 3),
  `Lower limit for number of crossings` = qbinom(0.05, n - 1, 0.5),
  check.names                           = FALSE)

knitr::kable(x)
```

