---
output: github_document
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  fig.height = 4,
  fig.width = 8,
  out.width = "100%",
  dev = 'png'
)

set.seed(1)
```

# Process Behaviour Charts

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

Run charts and individuals control charts for statistical quality control and improvement. 

`pbcharts` implements the I prime (or normalised I) control chart suggested by Taylor (2017) <https://variation.com/normalized-individuals-control-chart/>. The I prime chart adjusts the control limits to varying subgroup sizes making `pbcharts` useful for a wide range of measurement and count data and a convenient replacement for the classic Shewhart control charts.

`pbcharts` uses only functions from base R making in fast and robust.

* Facet plots (small multiples) on one categorical variable.

* Freeze calculations of centre line and control limits to a baseline period.

* Split charts into two periods.

* Exclude individual data points from calculations.

Please report any issues at https://github.com/anhoej/pbcharts/issues

## Installation

You can install the development version of pbcharts from [GitHub](https://github.com/anhoej/pbcharts) with:

```
devtools::install_github("anhoej/pbcharts")
```

## Examples

Draw a run chart of 24 random normal values:

```{r}
library(pbcharts)
pbc(rnorm(24))
```

Draw an individuals (I) control chart:

```{r}
pbc(rnorm(24), chart = 'i')
```

Signal special causes from data points outside control limits (red points) and(or) unusually long or few runs (red and dashed centre line):

```{r}
pbc(1:11, chart = 'i')
```

I chart of average decision to delivery times for grade 2 caesarian sections:

```{r}
pbc(month, avg_delay, 
data  = csection, 
chart = 'i')
```

Normalised I chart (or I prime chart) of C. section data taking varying subgroup sizes into account:

```{r}
pbc(month, avg_delay * n, n,  # multiply numerator and denominator to keep scale
    data = csection,
    chart = 'i')
```

Faceted I prime chart of bacteremia mortality in six hospitals:

```{r, fig.height=7}
pbc(month, deaths, cases,
    facet    = hospital,                # facet plot by hospital
    data     = bacteremia_mortality,
    chart    = 'i',
    multiply = 100,                     # show percent rather than proportions
    ylim     = c(0, NA),                # suppress negative control limits
    title    = 'Bacteremia mortality',
    ylab     = '%',
    xlab     = 'Month')
```

Print a summary:

```{r}
p <- pbc(month, deaths, cases, hospital,
         data  = bacteremia_mortality,
         chart = 'i',
         plot  = FALSE)

summary(p)
```

Plot a pbc object:

```{r, fig.height=7}
plot(p)
```

## Procedure for calculating centre line and conrol limits

We use the following symbols:

* n = numerators
* d = denominators
* o = number of data values
* i = i^th^ data value

Values to plot:

$$
y = \frac{n}{d}
$$

Centre line:

$$
CL = \frac{\sum{n}}{\sum{d}}
$$

Standard deviation of i^th^ data point:

$$
s_i = \sqrt{\frac{\pi}{2}}\frac{\vert{}y_i-y_{i-1}\vert{}}{\sqrt{\frac{1}{d_i}+\frac{1}{d_{i-1}}}}
$$

Average standard deviation:

$$
\bar{s} = \frac{\sum{s}}{o}
$$

Control limits:

$$
\text{control limits} = CL \pm 3 \frac{\bar{s}}{\sqrt{d_i}}
$$
