---
output:
  github_document:
    toc: true
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  fig.height = 4.5,
  fig.width = 9,
  out.width = "100%",
  dev = 'svg'
)
```

# Process Behaviour Charts

-- Yet another R package for statistical process control charts

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

An R package for run charts and individuals control charts for statistical quality control and improvement. 

pbcharts implements the I prime (I' or normalised I) control chart suggested by Taylor: <https://variation.com/normalized-individuals-control-chart/>.

The I' chart adjusts control limits to varying subgroup sizes making pbcharts useful for a wide range of measurement and count data and a convenient replacement for the classic Shewhart control charts.

Following the [tinyverse principles](https://www.tinyverse.org/), pbcharts relies solely on base R functions, avoiding external package dependencies, and making it both fast and robust.

pbcharts is currently able to:

* test for special cause variation using runs analysis and control limits;
* signal special causes using clear visual clues;
* freeze calculations of centre line and control limits to a baseline period;
* split charts into separate periods;
* exclude individual data points from calculations;
* facet plots (small multiples) on one or more categorical variables.

pbcharts is in early development. Please report any issues at https://github.com/anhoej/pbcharts/issues.

## Installation

You can install the development version of pbcharts from [GitHub](https://github.com/anhoej/pbcharts) with:

```
devtools::install_github("anhoej/pbcharts")
```

## Examples

### Basic usage

Draw a run chart of 24 random normal values:

```{r}
library(pbcharts)  # load pbcharts

set.seed(1)        # lock random number generator

y <- rnorm(24)     # generate random numbers to plot

pbc(y)             # plot run chart of 24 random normal values
```

Draw an individuals (I) control chart:

```{r}
pbc(y, chart = 'i')
```

Draw a moving standard deviations control chart:

```{r}
pbc(y, chart = 'ms')
```

Signal special causes from data points outside control limits (red points) and(or) unusually long or few runs (red and dashed centre line):

```{r}
pbc(1:11, chart = 'i')
```

See below for the rules applied to signal special cause variation.

### Plotting measurement data

Standard I chart (subgroup size = 1) of average decision to delivery times for grade 2 caesarian sections (C-section):

```{r}
pbc(month, avg_delay, 
    data  = csection, 
    chart = 'i')
```

I' chart of C-section data taking varying subgroup sizes into account:

```{r}
pbc(month, avg_delay * n, n,  # multiply numerator by denominator to keep scale
    data = csection,
    chart = 'i')
```

Notice that pbc() always plots the numerator divided by the denominator. Consequently, if the numerators have already been averaged, it is necessary to multiply the numerator by the denominator before plotting.

Standard I chart of average HbA1c in children with diabetes:

```{r}
pbc(month, avg_hba1c,
    data  = hba1c,
    chart = 'i',
    title = 'I chart of average HbA1c in children with diabetes',
    ylab  = 'mmol/mol',
    xlab  = 'Month')
```

I' chart of average HbA1c in children with diabetes using subgroups (number of children):

```{r}
pbc(month, avg_hba1c * n, n,
    data  = hba1c,
    chart = 'i',
    title = "I' chart of average HbA1c in children with diabetes",
    ylab  = 'mmol/mol',
    xlab  = 'Month')
```

### Plotting count data

Run chart of hospital infection rates:

```{r}
pbc(month, n, days,
    data     = cdi,
    multiply = 10000,
    title    = 'Hospital associated C. diff. infections',
    ylab     = 'Count per 10,000 risk days',
    xlab     = 'Month')
```

Freeze calculation of centre and control lines to period before intervention:

```{r}
pbc(month, n, days,
    data     = cdi,
    chart    = 'i',
    multiply = 10000,
    freeze   = 24,
    title    = 'Hospital associated C. diff. infections',
    ylab     = 'Count per 10,000 risk days',
    xlab     = 'Month')
```

Split chart after intervention:

```{r}
pbc(month, n, days,
    data     = cdi,
    chart    = 'i',
    multiply = 10000,
    split    = 24,
    title    = 'Hospital associated C. diff. infections',
    ylab     = 'Count per 10,000 risk days',
    xlab     = 'Month')
```

<!-- Ignore freak data point #20: -->

<!-- ```{r} -->
<!-- pbc(month, n, days, -->
<!--     data     = cdi, -->
<!--     chart    = 'i', -->
<!--     multiply = 10000, -->
<!--     split    = 24, -->
<!--     exclude  = 20, -->
<!--     title    = 'Hospital associated C. diff. infections', -->
<!--     ylab     = 'Count per 10,000 risk days', -->
<!--     xlab     = 'Month') -->
<!-- ``` -->

### Facetted charts for multidimensional data

Faceted I' chart of bacteremia mortality in six hospitals:

```{r, fig.height=7}
pbc(month, deaths, cases,
    facet    = hospital,                # facet plot by hospital
    data     = bacteremia_mortality,
    chart    = 'i',
    ypct     = TRUE,                    # show percentage rather than proportions
    title    = 'Bacteremia mortality',
    ylab     = '%',
    xlab     = 'Month')
```

Two-way faceted chart of average waiting times for 5 types of elective surgery in 5 regions.

```{r, fig.height=8}
pbc(qrt, avg_days,
    facet = list(region, operation), # facet by region and operation
    data  = waiting_times,
    title = 'Average waiting times for elective surgery',
    ylab  = 'Days',
    xlab  = 'Quarter')
```

### Structure and summary of a pbc object

Saving a pbc object

```{r}
# save pbc object while suppressing plotting
p <- pbc(month, deaths, cases,
         facet    = hospital,
         data     = bacteremia_mortality,
         chart    = 'i',
         ypct     = TRUE,
         title    = 'Bacteremia mortality',
         ylab     = '%',
         xlab     = 'Month',
         plot     = FALSE)             # Suppress plotting

# print structure
str(p)

# extract list element
p$title

# print summary
summary(p)

# plot chart
plot(p)
```

## Procedure for calculating centre line and control limits

We use the following symbols:

* n = numerators
* d = denominators
* o = number of data values
* i = i^th^ data value

Values to plot:

$$
y = \frac{n}{d}
$$

Centre line:

$$
CL = \frac{\sum{n}}{\sum{d}}
$$

Standard deviation of i^th^ data point:

$$
s_i = \sqrt{\frac{\pi}{2}}\frac{\vert{}y_i-y_{i-1}\vert{}}{\sqrt{\frac{1}{d_i}+\frac{1}{d_{i-1}}}}
$$

Average standard deviation:

$$
\bar{s} = \frac{\sum{s}}{o}
$$

Control limits:

$$
\text{control limits} = CL \pm 3 \frac{\bar{s}}{\sqrt{d_i}}
$$
## Tests for special cause variation

pbc() applies [three tests](https://doi.org/10.1186/s12874-018-0564-0) to detect special cause variation:

* **Data points outside the control limits**<br>
Any point falling outside the control limits indicates potential special cause variation.

* **Unusually long runs**<br>
A run consists of one or more consecutive data points on the same side of the centre line. Data points that fall exactly on the centre line neither contribute to nor interrupt a run. For a random process, the upper 95% prediction limit for the longest run is approximately `log₂(n) + 3`, rounded to the nearest whole number, where n is the number of useful data points (i.e. those not on the centre line).

* **Unusually few crossings**<br<
A crossing occurs when two consecutive data points fall on opposite sides of the centre line. In a random process, the number of crossings follows a binomial distribution. The lower 5% prediction limit can be found using the cumulative distribution function: `qbinom(p = 0.05, size = n - 1, prob = 0.5)`.

Data points outside the control limits are highlighted, and the centre line is dashed and coloured if either of the two runs tests is positive.

In run charts, the centre line represents the median; in control charts, it represents the (weighted) mean.

Critical values for runs and crossing:

```{r, results='asis', echo=FALSE}
n <- 10:100
x <- data.frame(
  `Number of useful observations`       = n,
  `Upper limit for longest run`         = round(log2(n) + 3),
  `Lower limit for number of crossings` = qbinom(0.05, n - 1, 0.5),
  check.names                           = FALSE)

knitr::kable(x)
```
